---
title: "Vectors"
subtitle: "[Reproducible data treatment with R](http://lmi.cnrs.fr/r/vectors.html)"
author: "Colin Bousige"
output: 
    learnr::tutorial:
        progressive: true
        allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
# rmarkdown::run("vectors.Rmd")
library(learnr)
library(tidyverse)
gradethis::gradethis_setup()
```

## Exercise 1

### 

Consider two vectors `x` and `y` such as:
```{r}
x <- 1:5
y <- seq(0, 4, along=x)
```

Without typing it into Rstudio, what are the values of `x`, `y`, and `x*y`?
```{r q0-setup, include=FALSE}
x <- 1:5
y <- seq(0, 4, along = x)
```

```{r q0, exercise=TRUE, exercise.blanks = "___+"}

```


###

Consider two vectors `a` and `b` such as:

```{r}
a <- c(1,5,4,3,6)
b <- c(3,5,2,1,9)
```
Without typing it into Rstudio, what is the value of: `a<=b`{.R}?

```{r q1, exercise=TRUE, exercise.blanks = "___+"}
a <- c(1, 5, 4, 3, 6)
b <- c(3, 5, 2, 1, 9)
___
```


###

If 
```{r}
x <- c(1:12, NA, 5:2)
```
Without typing it into Rstudio, what is the value of: `length(x)`{.R}

```{r q2, exercise=TRUE, exercise.blanks = "___+"}
x <- ___
___
```

###

If 
```{r}
a <- 12:5
```
Without typing it into Rstudio, what is the value of: `is.numeric(a)`{.R}

```{r q3, exercise=TRUE, exercise.blanks = "___+"}
a <- 12:5
is.numeric(a)
```

###

Consider two vectors `x` and `y` such as:
```{r}
x <- 12:4
y <- c(0, 1, 2, 0, 1, 2, 0, 1, 2)
```
Without typing it into Rstudio, what is the value of: `which(!is.finite(x/y))`{.R}?

```{r q4, exercise=TRUE, exercise.blanks = "___+"}
x <- 12:4
y <- c(0, 1, 2, 0, 1, 2, 0, 1, 2)
___
```

###

If 
```{r}
x <- c('blue', 'red', 'green', 'yellow')
```
Without typing it into Rstudio, what is the value of: `is.character(x)`{.R}?

```{r q5, exercise=TRUE, exercise.blanks = "___+"}
x <- ___
___
```

###

If 
```{r}
x <- c('blue', 10, 'green', 20)
```
Without typing it into Rstudio, what is the value of: `is.character(x)`{.R}?

```{r q6, exercise=TRUE, exercise.blanks = "___+"}
x <- ___
___
```

###

Assign value 5 to the variable x.

Is there a difference between `1:x-1`{.R} and `1:(x-1)`{.R} ?

Explain.

```{r q7, exercise=TRUE, exercise.blanks = "___+"}
___
___
___
```

### 

Generate the sequence `9, 18, 27, 36, 45, 54, 63, 72, 81, 90`{.R} in 4 different manners.

```{r q8, exercise=TRUE, exercise.blanks = "___+"}
c(___)
seq(___)
___:___
___
```

```{r q8-solution, exercise.reveal_solution = TRUE}
c(9, 18, 27, 36, 45, 54, 63, 72, 81, 90)
seq(9, 90, by = 9)
seq(9, 90, length = 10)
seq(1, 10, 1)*9
1:10*9
```


###

```{r quizz, echo=FALSE}
quiz(
question("If `x <- c(\"w\", \"h\", \"f\", \"g\", \"k\")`, what will be the output for `x[c(2,3)]`?",
  answer("\"h\", \"f\"", correct = TRUE),
  answer("\"h\""),
  answer("\"f\""),
  incorrect = "Incorrect!"
),
question("Without typing it into Rstudio, if `x <- c(\"w\", \"h\", \"f\", \"g\", \"k\")`, what will be the third value in the index vector operation `x[c(2, 4, 4)]`?",
  answer("\"h\""),
  answer("NA"),
  answer("\"g\"", correct = TRUE),
  incorrect = "Incorrect!"
),
question("Without typing it into Rstudio, if `x <- c(\"w\", \"h\", \"f\", \"g\", \"k\")`, what will be the fourth value in the index vector operation `x[-2]`",
  answer("\"h\""),
  answer("\"g\""),
  answer("\"k\"", correct = TRUE),
  incorrect = "Incorrect!"
)
)
```


###

Let `a <- c(2, 4, 6, 8)`{.R} and `b <- c(TRUE, FALSE, TRUE, FALSE)`{.R}.

Without typing it into Rstudio, what will be the output for the R expression `max(a[b])`{.R}?

```{r q9, exercise=TRUE, exercise.blanks = "___+"}

```

## Exercice 2


We have the following `x` vector: `x <- c("10K", "100K", "200K", "500K", "1000K")`. Get rid of the  `"K"` character and turn this into a numerical vector.

```{r q10-setup, include=FALSE}
x <- c("10K", "100K", "200K", "500K", "1000K")
```


```{r q10, exercise=TRUE, exercise.blanks = "___+"}
x
```

```{r q10-hint}
# Use gsub() and as.numeric()
```

```{r q10-solution, exercise.reveal_solution = FALSE}
x <- as.numeric(gsub("K", "", x))
```

```{r q10-code-check}
grade_code()
```


## Exercice 3

We have the following `times` vector: `times <- c("010_min", "100_sec", "200_sec", "050_min")`. We want a numerical vector containing times in seconds only.

###

Using `substr()`, create 2 vectors `times_values` and `times_units` containing the numbers and the units from `times`.

```{r exo3-1-setup, include=FALSE}
times <- c("010_min", "100_sec", "200_sec", "050_min")
```


```{r exo3-1, exercise=TRUE, exercise.blanks = "___+"}
times_values <- substr(string, start, stop)
times_units  <- ___
```

```{r exo3-1-solution, exercise.reveal_solution = FALSE}
times_values <- as.numeric(substr(times, 1, 3))
times_units  <- substr(times, 5, 7)
```

```{r exo3-1-code-check}
grade_code()
```


###

You could do the same using `strsplit()` and `unlist()`

```{r exo3-2-setup, include=FALSE}
times <- c("010_min", "100_sec", "200_sec", "050_min")
```


```{r exo3-2, exercise=TRUE, exercise.blanks = "___+"}
strsplit("test", "e")
unlist(strsplit("test", "e"))
```

```{r exo3-2-solution, exercise.reveal_solution = FALSE}
times_values <- as.numeric(unlist(strsplit(times, "_"))[1:8 %% 2 == TRUE])
times_units  <- unlist(strsplit(times, "_"))[1:8%%2==FALSE]
```

```{r exo3-2-code-check}
grade_code()
```


###

Now, using `ifelse(test, yes, no)`, create the `times_sec` vector containing the numerical values of time all converted to seconds.

```{r exo3-3-setup, include=FALSE}
times <- c("010_min", "100_sec", "200_sec", "050_min")
times_values <- as.numeric(unlist(strsplit(times, "_"))[1:8 %% 2 == TRUE])
times_units <- unlist(strsplit(times, "_"))[1:8 %% 2 == FALSE]
```


```{r exo3-3, exercise=TRUE, exercise.blanks = "___+"}
times_sec <- ifelse(test, yes, no)
```

```{r exo3-3-solution, exercise.reveal_solution = TRUE}
times_sec <- ifelse(times_units == "min", times_values*60, times_values)
times_sec
```

```{r exo3-3-code-check}
grade_code()
```


## Exercise 4

Let's create a named vector containing age of the students in the class, the names of each value being the first name of the students. 

```{r exo2-1, exercise=TRUE, exercise.blanks = "___+"}
students <- c(
    "John" = 35,
    "Jane" = 36
)
students
```

Then (do it in Rstudio this time):

- Compute the average age and its standard deviation
- Compute the median age
- What is the maximum, minimum and range of the ages in the class?
- What are all the student names in the class?
- Print the sorted ages by increasing and decreasing order
- Print the ages sorted by alphabetically ordered names (increasing and decreasing)
- Show a histogram of the ages distribution using `hist()`{.R} and play with the parameter `breaks` to modify the histogram
- Show a boxplot of the ages distribution using `boxplot()`{.R}


## Exercise 5

Let's say we have this population data for these French cities for the two years 1962 and 2012:
```{r exo5-setup, echo=FALSE}
cities <- c(
    "Angers", "Bordeaux", "Brest", "Dijon", "Grenoble", "Le Havre",
    "Le Mans", "Lille", "Lyon", "Marseille", "Montpellier", "Nantes",
    "Nice", "Paris", "Reims", "Rennes", "Saint-Etienne", "Strasbourg",
    "Toulon", "Toulouse"
)
pop_1962 <- c(
    115273, 278403, 136104, 135694, 156707, 187845, 132181, 239955,
    535746, 778071, 118864, 240048, 292958, 2790091, 134856, 151948,
    210311, 228971, 161797, 323724
)
pop_2012 <- c(
    149017, 241287, 139676, 152071, 158346, 173142, 143599, 228652,
    496343, 852516, 268456, 291604, 343629, 2240621, 181893, 209860,
    171483, 274394, 164899, 453317
)
df <- tibble::tibble(cities, "1962" = pop_1962, "2012" = pop_2012) %>%
    pivot_longer(cols = -cities, names_to = "year", values_to = "pop") %>%
    pivot_wider(names_from = "cities", values_from = "pop")
knitr::kable(df[,1:10])
knitr::kable(df[,11:ncol(df)])
```

###

Create a `cities` vector containing all the cities listed above:

```{r exo5-1, exercise=TRUE, exercise.blanks = "___+", exercise.setup="exo5-setup"}
cities <- c(___)
```

```{r exo5-1-solution, exercise.reveal_solution = FALSE}
cities <- c(
    "Angers", "Bordeaux", "Brest", "Dijon", "Grenoble", "Le Havre",
    "Le Mans", "Lille", "Lyon", "Marseille", "Montpellier", "Nantes",
    "Nice", "Paris", "Reims", "Rennes", "Saint-Etienne", "Strasbourg",
    "Toulon", "Toulouse"
)
```

```{r exo5-1-code-check}
grade_code()
```

###

Create a `pop_1962` and `pop_2012` vectors containing the populations of each city at these years. Print the 2 vectors. 

```{r exo5-2, exercise=TRUE, exercise.blanks = "___+", exercise.setup="exo5-setup"}
pop_1962 <- c(___)
pop_2012 <- c(___)
```

```{r exo5-2-solution, exercise.reveal_solution = FALSE}
pop_1962 <- c(
    115273, 278403, 136104, 135694, 156707, 187845, 132181, 239955,
    535746, 778071, 118864, 240048, 292958, 2790091, 134856, 151948,
    210311, 228971, 161797, 323724
)
pop_2012 <- c(
    149017, 241287, 139676, 152071, 158346, 173142, 143599, 228652,
    496343, 852516, 268456, 291604, 343629, 2240621, 181893, 209860,
    171483, 274394, 164899, 453317
)
```

```{r exo5-2-code-check}
grade_code()
```

###

Use `names()`{.R} to name values of `pop_1962` and `pop_2012`. Print the 2 vectors again. Are there any change?

```{r exo5-3, exercise=TRUE, exercise.blanks = "___+", exercise.setup="exo5-setup"}
names(pop_1962) <- ___
names(pop_2012) <- ___
```


```{r exo5-3-solution, exercise.reveal_solution = FALSE}
names(pop_1962) <- cities
names(pop_2012) <- cities
pop_1962
pop_2012
```

```{r exo5-3-code-check}
grade_code()
```

```{r exo5-setup2}
cities <- c(
    "Angers", "Bordeaux", "Brest", "Dijon", "Grenoble", "Le Havre",
    "Le Mans", "Lille", "Lyon", "Marseille", "Montpellier", "Nantes",
    "Nice", "Paris", "Reims", "Rennes", "Saint-Etienne", "Strasbourg",
    "Toulon", "Toulouse"
)
pop_1962 <- c(
    115273, 278403, 136104, 135694, 156707, 187845, 132181, 239955,
    535746, 778071, 118864, 240048, 292958, 2790091, 134856, 151948,
    210311, 228971, 161797, 323724
)
pop_2012 <- c(
    149017, 241287, 139676, 152071, 158346, 173142, 143599, 228652,
    496343, 852516, 268456, 291604, 343629, 2240621, 181893, 209860,
    171483, 274394, 164899, 453317
)
names(pop_2012) <- names(pop_1962) <- cities
```

###

What are the cities with more than 200000 people in 1962? For these, how many residents in 2012?

```{r exo5-4, exercise=TRUE, exercise.blanks = "___+", exercise.setup="exo5-setup2"}
pop_1962 > 200000
```

```{r exo5-4-solution, exercise.reveal_solution = TRUE}
cities200k <- cities[pop_1962 > 200000]
cities200k
pop_2012[cities200k]
```

```{r exo5-4-code-check}
grade_code()
```


### 

What is the population evolution of Montpellier and Nantes?

```{r exo5-5, exercise=TRUE, exercise.blanks = "___+", exercise.setup="exo5-setup2"}
pop_2012 - pop_1962
```

```{r exo5-5-solution, exercise.reveal_solution = TRUE}
pop_2012["Montpellier"] - pop_1962["Montpellier"]
pop_2012["Nantes"] - pop_1962["Nantes"]
```

```{r exo5-5-code-check}
grade_code()
```

### 

Create a `pop_diff` vector to store population change between 1962 and 2012

```{r exo5-6, exercise=TRUE, exercise.blanks = "___+", exercise.setup="exo5-setup2"}
pop_diff <- ___
```

```{r exo5-6-solution, exercise.reveal_solution = TRUE}
pop_diff <- pop_2012 - pop_1962
```

```{r exo5-6-code-check}
grade_code()
```

### 

```{r exo5-setup3}
cities <- c(
    "Angers", "Bordeaux", "Brest", "Dijon", "Grenoble", "Le Havre",
    "Le Mans", "Lille", "Lyon", "Marseille", "Montpellier", "Nantes",
    "Nice", "Paris", "Reims", "Rennes", "Saint-Etienne", "Strasbourg",
    "Toulon", "Toulouse"
)
pop_1962 <- c(
    115273, 278403, 136104, 135694, 156707, 187845, 132181, 239955,
    535746, 778071, 118864, 240048, 292958, 2790091, 134856, 151948,
    210311, 228971, 161797, 323724
)
pop_2012 <- c(
    149017, 241287, 139676, 152071, 158346, 173142, 143599, 228652,
    496343, 852516, 268456, 291604, 343629, 2240621, 181893, 209860,
    171483, 274394, 164899, 453317
)
names(pop_2012) <- names(pop_1962) <- cities
pop_diff <- pop_2012 - pop_1962
```

Print cities with a negative change

```{r exo5-7, exercise=TRUE, exercise.blanks = "___+", exercise.setup="exo5-setup3"}
pop_diff
```

```{r exo5-7-solution, exercise.reveal_solution = TRUE}
cities[pop_diff < 0]
```

```{r exo5-7-code-check}
grade_code()
```


### 

Print cities which broke the 300000 people barrier between 1962 and 2012

```{r exo5-8, exercise=TRUE, exercise.blanks = "___+", exercise.setup="exo5-setup3"}
pop_2012 > 300000
```

```{r exo5-8-solution, exercise.reveal_solution = TRUE}
cities[pop_2012 > 300000 & pop_1962 < 300000]
```

```{r exo5-8-code-check}
grade_code()
```

### 

Compute the total change in population of the 10 largest cities (as of 1962) between 1962 and 2012.

```{r exo5-9, exercise=TRUE, exercise.blanks = "___+", exercise.setup="exo5-setup3"}
order(pop_1962)
ten_largest <- cities[___]
sum(___)
```

```{r exo5-9-solution, exercise.reveal_solution = TRUE}
ten_largest <- cities[order(pop_1962, decreasing = TRUE)[1:10]]
sum(pop_2012[ten_largest] - pop_1962[ten_largest])
```

```{r exo5-9-code-check}
grade_code()
```

### 

Compute the population mean for year 1962

```{r exo5-10, exercise=TRUE, exercise.blanks = "___+", exercise.setup="exo5-setup3"}

```

```{r exo5-10-solution, exercise.reveal_solution = TRUE}
mean(pop_1962)
```

```{r exo5-10-code-check}
grade_code()
```


### 

Compute the population mean of Paris over these two years

```{r exo5-11, exercise=TRUE, exercise.blanks = "___+", exercise.setup="exo5-setup3"}
mean(___)
```

```{r exo5-11-solution, exercise.reveal_solution = TRUE}
mean(c(pop_1962['Paris'], pop_2012['Paris']))
```

```{r exo5-11-code-check}
grade_code()
```

### 

Sort the cities by decreasing order of population for 1962

```{r exo5-12, exercise=TRUE, exercise.blanks = "___+", exercise.setup="exo5-setup3"}

```

```{r exo5-12-solution, exercise.reveal_solution = TRUE}
sort(pop_1962, decreasing = TRUE)
```

```{r exo5-12-code-check}
grade_code()
```
