---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tutoR)
```

# Description

This package is a gathering of various interactive exercises for the "[Reproducible data treatment with R](http://lmi.cnrs.fr/r/)" class.

It also contains some user-defined functions that are used in the class.

# Usage

## Tutorials

After having installed the package, launch a tutorial by running:

```r
learnr::run_tutorial("tutorial_name", package = "tutoR")
```

In this call, `"tutorial_name"` can be any of:

- `"variables"`
- `"vectors"`
- `"dataframes"`
- `"plots"`
- `"fits"`

## Example Files

The `tutoR` package also come bundled with some example files. A complete list can be accessed with:

```r
tutoR_example()
tutoR_example(pattern = "csv") # apply a regex filter
# To read a file, for example:
read.csv(tutoR_example("CO2_emission.csv"))
```

## Functions

List of functions included in the package:

- [`dxdy()`](R/dxdy.R)
- [`Gaussian()`](R/Gaussian.R)
- [`integ()`](R/integ.R)
- [`Lorentzian()`](R/Lorentzian.R)
- [`norm01()`](R/norm01.R)
- [`Pruby()`](R/Pruby.R)
- [`Rshift_to_lambda()`](R/Rshift_to_lambda.R)
- [`sinc()`](R/sinc.R)
- [`tutoR_examples()`](R/tutoR_example.R)
