---
title: "Variables"
subtitle: "[Reproducible data treatment with R](http://lmi.cnrs.fr/r/variables-booleans-and-strings.html)"
author: Colin Bousige
output: 
    learnr::tutorial:
        progressive: true
        allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
# rmarkdown::run("variables.Rmd")
library(learnr)
gradethis::gradethis_setup()
```

## Variables

###

This code computes the answer to one plus one, change it so it computes two plus two:

```{r addition, exercise=TRUE}
1 + 1
```

###

Attribute to `x` the value `2.5`, and compute the exponential of its squared value:

```{r qblank, exercise=TRUE, exercise.blanks = "___+"}
x <- ___
```


```{r qblank-solution, exercise.reveal_solution = FALSE}
x <- 2.5
exp(x^2)
```

```{r qblank-code-check}
grade_code()
```

###

Attribute the string `"100.3"` to the variable `x`, and the value 99.6 to the variable `y`:
```{r q0, exercise=TRUE, exercise.blanks = "___+"}
x <- ___
y <- ___
```

```{r q0-solution, exercise.reveal_solution = FALSE}
x <- "100.3"
y <- 99.6
```

```{r q0-code-check}
grade_code()
```

###

Assuming the `x` and `y` defined above, modify the following code so that it does not return an error:

```{r q1-setup, include=FALSE}
x <- "100.3"
y <- 99.6
```

```{r q1, exercise=TRUE}
floor(x) - ceiling(y)
```

```{r q1-hint}
# look into 
as.numeric()
```

```{r q1-solution, exercise.reveal_solution = FALSE}
floor(as.numeric(x)) - ceiling(y)
```

```{r q1-code-check}
grade_code()
```


## Booleans

###

I defined two random variables `x` and `y`. Are the two variables equal? which one is smaller?

```{r qtest-setup, include=FALSE}
x <- runif(1)
y <- runif(1)
```

```{r qtest, exercise=TRUE, exercise.blanks = "___+"}

```

```{r qtest-solution, exercise.reveal_solution = FALSE}
x == y
```

```{r qtest-code-check}
grade_code()
```

###

I defined two random variables `x` and `y`. Using the `ifelse()`{.R} function, print "x is larger than y" or "x is smaller than y".

```{r qtest2-setup, include=FALSE}
x <- runif(1)
y <- runif(1)
```

```{r qtest2, exercise=TRUE, exercise.blanks = "___+"}
ifelse(test, yes, no)
```

```{r qtest2-solution, exercise.reveal_solution = FALSE}
ifelse(x < y, "x is smaller than y", "x is larger than y")
```

```{r qtest2-code-check}
grade_code()
```


### 

Without running the code, what do you think will be the output of this code:
```{r q2, exercise=TRUE}
cos(pi/2)==cos(3*pi/2)
```

```{r q3, exercise=TRUE}
cos(pi/2)-cos(3*pi/2)
```

In fact, `.Machine$double.eps` is the smallest positive floating-point number x such that `1 + x != 1`. It depends on the machine (computer) you are running on:
```{r machine, exercise=TRUE}
.Machine$double.eps
(cos(pi / 2) - cos(3 * pi / 2)) > .Machine$double.eps
all.equal(cos(pi/2),cos(3*pi/2))
```


## Strings

Save your name as a string in a variable, and print "My name is: yourname" using this variable.

```{r qstring1, exercise=TRUE, exercise.blanks = "___+"}
name <- ___
```

```{r qstring1-hint}
name <- "___+"
```

```{r qstring1-solution, exercise.reveal_solution = FALSE}
name <- "Colin"
paste("My name is:", name)
```

```{r qstring1-code-check}
grade_code()
```

###

Replace all "e" by "A" in the following string:

```{r qstring2, exercise=TRUE, exercise.blanks = "___+"}
x <- "aaaeebbbeebiieelakdceee"
gsub(___, ___, ___)
```

```{r qstring2-solution, exercise.reveal_solution = FALSE}
x <- "aaaeebbbeebiieelakdceee"
gsub("e", "A", x)
```

```{r qstring2-code-check}
grade_code()
```

###

Do the same using the `stringr` function `str_replace_all()`{.R} (take a look at the [cheatsheet](https://github.com/rstudio/cheatsheets/blob/main/strings.pdf)):

```{r qstringr, exercise=TRUE, exercise.blanks = "___+"}
library(stringr)
x <- "aaaeebbbeebiieelakdceee"
str_replace_all(___, ___, ___)
```

```{r qstringr-solution, exercise.reveal_solution = FALSE}
library(stringr)
x <- "aaaeebbbeebiieelakdceee"
str_replace_all(x, "e", "A")
```

```{r qstringr-code-check}
grade_code()
```


###

Find the index of all the `"e"` characters in the string, using base functions and `stringr` ones (take a look at the [cheatsheet](https://github.com/rstudio/cheatsheets/blob/main/strings.pdf) to find the good one):

```{r qstringr2, exercise=TRUE, exercise.blanks = "___+"}
x <- "aaaeebbbeebiieelakdceee"
gregexpr(___, ___)
library(stringr)
str____(___, ___)
```

```{r qstringr2-solution, exercise.reveal_solution = FALSE}
x <- "aaaeebbbeebiieelakdceee"
gregexpr("e", x)
library(stringr)
str_locate_all(x, "e")
```

```{r qstringr2-code-check}
grade_code()
```
