---
title: "Data Frames"
subtitle: "[Reproducible data treatment with R](http://lmi.cnrs.fr/r/data-frames.html)"
author: "Colin Bousige"
output: 
    learnr::tutorial:
        progressive: true
        allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
# rmarkdown::run("dataframe.Rmd")
library(learnr)
library(tutor)
library(tidyverse)
gradethis::gradethis_setup()
```

## Exercise 1

###

Create a 3 column `data.frame`{.R} called `df` containing three columns `x`, `y` and `z` with:

- `x` a vector from $-\pi$ to $\pi$ of length 10
- `y` their sinus
- and `z` the sum of the two first columns.

```{r q0, exercise=TRUE, exercise.blanks = "___+"}
x <- ___
y <- ___
z <- ___
df <- data.frame()
```

```{r q0-solution, exercise.reveal_solution = FALSE}
x <- seq(-pi, pi, length=10)
y <- sin(x)
z <- x + y
df <- data.frame(x=x, y=y, z=z)
```

```{r q0-code-check}
grade_code()
```
###

Print the 4 first lines of the table.

```{r q1, exercise=TRUE, exercise.blanks = "___+", exercise.setup="q0-solution"}

```

```{r q1-hint}
# Take a look at the head() function
```

```{r q1-solution, exercise.reveal_solution = FALSE}
head(df, 4)
```

```{r q1-code-check}
grade_code()
```

###

Print the second (*i.e.* `y`) column with three different ways.

```{r q2, exercise=TRUE, exercise.blanks = "___+", exercise.setup="q0-solution"}
df[___]
df[___]
df$___
```

```{r q2-solution, exercise.reveal_solution = FALSE}
df[,2]
df[,"y"]
df$y
```

```{r q2-code-check}
grade_code()
```


###

Modify the column `z` so that it contains its value minus its minimum.

```{r q2bis, exercise=TRUE, exercise.blanks = "___+", exercise.setup="q0-solution"}

```

```{r q2bis-solution, exercise.reveal_solution = FALSE}
df$z <- df$z - min(df$z)
```

```{r q2bis-code-check}
grade_code()
```

###

Print the average of the third column.

```{r q3, exercise=TRUE, exercise.blanks = "___+", exercise.setup="q0-solution"}

```


```{r q3-hint}
# look at the mean() function
```

```{r q3-solution, exercise.reveal_solution = FALSE}
mean(df$z)
```

```{r q3-code-check}
grade_code()
```

###

Using `plot(x,y)`{.R} where `x` and `y` are vectors, plot the 2nd column as a function of the first.


```{r q4, exercise=TRUE, exercise.blanks = "___+", exercise.setup="q0-solution"}
plot(___, ___)
```


```{r q4-solution, exercise.reveal_solution = FALSE}
plot(df$x, df$y)
```

```{r q4-code-check}
grade_code()
```

###

Look into the function `write.table()`{.R} to write a text file containing this `data.frame`{.R}.

```{r q5, exercise=TRUE, exercise.blanks = "___+", exercise.setup="q0-solution"}
write.table(___)
```


```{r q5-solution, exercise.reveal_solution = FALSE}
write.table(df, "~/Downloads/some_data.dat", quote = FALSE, row.names = FALSE)
```

```{r q5-code-check}
grade_code()
```


###

Do the all the same things with a `tibble`{.R}.

```{r q6, exercise=TRUE, exercise.blanks = "___+"}
df_tib <- tibble(___)
```

```{r q6-solution, exercise.reveal_solution = TRUE}
library(tidyverse)
df_tib <- tibble(x = seq(-pi, pi, length = 10), 
                 y = sin(x), 
                 z = x + y)
head(df_tib, 4)
df_tib[, 2]
df_tib[[2]]
mean(df_tib$z)
write_csv(df_tib, "~/Downloads/some_data.csv")
plot(df_tib$x, df_tib$y)
```

```{r q6-code-check}
grade_code()
```


## Exercise 2

We will work with 3 different files:
    - `tutor_examples("rubis_01.txt")`
    - `tutor_examples("population.csv")`
    - `tutor_examples("FTIR_rocks.xlsx")`


###

Load them into separate `data.frames`{.R}. Look into the options of `read.table()`{.R}, `read.csv()`{.R}, `readxl::read_excel()`{.R}, to get the proper data fields. Make sure that the `rubis_01` data.frame has `w` and `intensit` as column names.

```{r exo2-1, exercise=TRUE, exercise.blanks = "___+"}
rubis_01   <- ___(tutor_examples("rubis_01.txt"), ___)
population <- ___(tutor_examples("population.csv"))
FTIR_rocks <- ___(tutor_examples("FTIR_rocks.xlsx"))
```

```{r exo2-1-solution, exercise.reveal_solution = FALSE}
rubis_01   <- read.table(tutor_examples("rubis_01.txt"), col.names = c("w", "intensity"))
population <- read.csv(tutor_examples("population.csv"))
FTIR_rocks <- readxl::read_excel(tutor_examples("FTIR_rocks.xlsx"))
```

```{r exo2-1-code-check}
grade_code()
```


### 

Print their dimensions and column names. 

```{r exo2-2, exercise=TRUE, exercise.blanks = "___+", exercise.setup="exo2-1-solution"}
# Dimensions
rubis_01
population
FTIR_rocks
# Names
rubis_01
population
FTIR_rocks
```

```{r exo2-2-solution, exercise.reveal_solution = FALSE}
dim(rubis_01);   names(rubis_01)
dim(population); names(population)
dim(FTIR_rocks); names(FTIR_rocks)
```

```{r exo2-2-code-check}
grade_code()
```

###

Do the same things by loading directly into tibbles.

```{r exo2-3, exercise=TRUE, exercise.blanks = "___+", exercise.setup="exo2-1-solution"}

```

```{r exo2-3-solution, exercise.reveal_solution = FALSE}
library(tidyverse)
rubis_01 <- read_table(tutor_examples("rubis_01.txt"), col_names = c("w", "intensity"))
population <- read_csv(tutor_examples("population.csv"))
```

```{r exo2-3-code-check}
grade_code()
```


## Exercise 3

We will use the TGA data file `tutor_examples("ATG.txt")` ([click link](https://github.com/colinbousige/tutor/blob/master/inst/extdata/ATG.txt) to take a look at the file).

Load it into a `data.frame`{.R}. Look into the options of [`read.table()`{.R}](https://www.rdocumentation.org/packages/utils/versions/3.6.2/topics/read.table) to get the proper data fields.

```{r exo3-1, exercise=TRUE, exercise.blanks = "___+"}
d <- read.table(tutor_examples("ATG.txt"),
                ___
                )
d
```

```{r exo3-1-hint}
# check how many lines you have to read
# check how many lines you have to skip before reading
# you need to skip the line with the unit
```

```{r exo3-1-solution, exercise.reveal_solution = TRUE}
# Two versions
d <- read.table(tutor_examples("ATG.txt"),
    skip = 12,
    header = FALSE,
    nrows = 4088
)
names(d) <- c("Index", "t", "Ts", "Tr", "Value")
head(d)
d <- read.table(tutor_examples("ATG.txt"),
    skip = 10,
    comment.char = "[",
    header = TRUE,
    nrows = 4088
)
head(d)
```

###

Do the same using the `tidyverse` function [`read_table()`](https://www.rdocumentation.org/packages/readr/versions/1.3.1/topics/read_table):

```{r exo3-2, exercise=TRUE, exercise.blanks = "___+"}
library(tidyverse)
d <- read_table(tutor_examples("ATG.txt"),
                ___
                )
d
```

```{r exo3-2-solution, exercise.reveal_solution = TRUE}
library(tidyverse)
d <- read_table(tutor_examples("ATG.txt"), 
                skip    = 10,
                comment = "[") %>% 
        drop_na()
d
```


## Exercise 4

### 

Load the `tutor_examples("population.csv")` file into a `tibble`{.R} called `popul`.

```{r exo4-1, exercise=TRUE, exercise.blanks = "___+"}
library(___)
popul <- read____(tutor_examples("population.csv"))
```

```{r exo4-1-solution, exercise.reveal_solution = FALSE}
library(tidyverse)
popul <- read_csv(tutor_examples("population.csv"))
```

```{r exo4-1-code-check}
grade_code()
```

###

What are the names of the columns? What's the dimension of the table ?

```{r exo4-2, exercise=TRUE, exercise.blanks = "___+", exercise.setup="exo4-1-solution"}
popul
```

```{r exo4-2-solution, exercise.reveal_solution = FALSE}
names(popul); dim(popul)
```

```{r exo4-2-code-check}
grade_code()
```

###

Are the data tidy? make the table tidy if needed

```{r exo4-3, exercise=TRUE, exercise.blanks = "___+", exercise.setup="exo4-1-solution"}
popul
popul.tidy <- popul %>% 
    pivot_longer(
        cols     = ___, # what are the columns we want to keep? -> -these
        names_to = ___, # name of the column gathering the original column names
        values_to= ___  # name of the column gathering the original column values
        )
```

```{r exo4-3-solution, exercise.reveal_solution = TRUE}
head(popul) # no
popul.tidy <- popul %>% 
    pivot_longer(cols      = -year,
                 names_to  = "city",
                 values_to = "pop"
                )
popul.tidy
```

```{r exo4-3-code-check}
grade_code()
```


###

Create a subset containing the data for Montpellier using a [filtering](https://dplyr.tidyverse.org/reference/filter.html) function from the `tidyverse`.

```{r exo4-4, exercise=TRUE, exercise.blanks = "___+", exercise.setup="exo4-3-solution"}
mtp <- popul.tidy %>% ___
```

```{r exo4-4-solution, exercise.reveal_solution = FALSE}
mtp <- popul.tidy %>% filter(city == "Montpellier")
```

```{r exo4-4-code-check}
grade_code()
```

- What is the max and min of population in this city?
- The average population over time?

```{r exo4-5, exercise=TRUE, exercise.blanks = "___+", exercise.setup="exo4-4-solution"}
mtp
```

```{r exo4-5-solution, exercise.reveal_solution = FALSE}
max(mtp$pop)
min(mtp$pop)
range(mtp$pop)
mean(mtp$pop)
```

```{r exo4-5-code-check}
grade_code()
```

###

What is the total population in 2012?

```{r exo4-6, exercise=TRUE, exercise.blanks = "___+", exercise.setup="exo4-4-solution"}
popul.tidy %>% 
    ___ %>%         # You need to filter the data for the year 2012
    ___ %>%         # Then select the right column
    ___             # And perform the sum of its data
```

```{r exo4-6-solution, exercise.reveal_solution = TRUE}
sum(popul.tidy[popul.tidy$year == 2012, "pop"])
popul.tidy %>%
    filter(year == 2012) %>%
    select(pop) %>%
    sum()
```

```{r exo4-6-code-check}
grade_code()
```

###

What is the total population per year?

```{r exo4-7, exercise=TRUE, exercise.blanks = "___+", exercise.setup="exo4-4-solution"}
popul.tidy %>% 
    ___ %>%    # You need to group data per year
    ___        # Then summarize the data of each year as 
               # the total population of each group
```

```{r exo4-7-solution, exercise.reveal_solution = FALSE}
popul.tidy %>% 
    group_by(year) %>% 
    summarise(pop_tot = sum(pop))
```

```{r exo4-7-code-check}
grade_code()
```

###

What is the average population per city over the years?

```{r exo4-8, exercise=TRUE, exercise.blanks = "___+", exercise.setup="exo4-4-solution"}
popul.tidy %>%
    ___ %>%  # You need to group data per...?
    ___      # Then...?
```

```{r exo4-8-solution, exercise.reveal_solution = FALSE}
popul.tidy %>%
    group_by(city) %>%
    summarise(pop_ave = mean(pop))
```

```{r exo4-8-code-check}
grade_code()
```


## Exercise 5

First, load the `tidyverse` and `lubridate` package

```{r exo5-1, exercise=TRUE, exercise.blanks = "___+"}
___
___
```


```{r exo5-1-solution, exercise.reveal_solution = FALSE}
library(tidyverse)
library(lubridate)
```

```{r exo5-1-code-check}
grade_code()
```

###

Load `tutor_examples("people1.csv")` and `tutor_examples("people2.csv")` into `pp1` and `pp2`, and take a look at them.

```{r exo5-2, exercise=TRUE, exercise.blanks = "___+", exercise.setup="exo5-1-solution"}
tutor_examples("people1.csv")
tutor_examples("people2.csv")
pp1 <- read____(___)
pp2 <- read____(___)
```

```{r exo5-2-solution, exercise.reveal_solution = FALSE}
pp1 <- read_csv(tutor_examples("people1.csv"))
pp2 <- read_csv(tutor_examples("people2.csv"))
```

```{r exo5-2-code-check}
grade_code()
```


### 

Create a new tibble `pp` by using the pipe operator (`%>%`{.R}) and successively:

- joining the two tibbles into one using `inner_join()`{.R}
- adding a column `age` containing the age in years (use `lubridate::time_length(x, 'years')`{.R} with x a time difference in days) by using `mutate()`{.R}

```{r exo5-3, exercise=TRUE, exercise.blanks = "___+", exercise.setup="exo5-2-solution"}
pp <- pp1 %>%
    ___ %>%       # you need to join with pp2
    mutate(___)   # then add a column `age` computing the right thing
```

```{r exo5-3-solution, exercise.reveal_solution = TRUE}
pp <- pp1 %>%
    inner_join(pp2) %>%
    mutate(age = time_length(today() - dateofbirth, "years"))
```

```{r exo5-3-code-check}
grade_code()
```

###

Display a summary of the table using `str()`{.R}

```{r exo5-4, exercise=TRUE, exercise.blanks = "___+", exercise.setup="exo5-3-solution"}

```

```{r exo5-4-solution, exercise.reveal_solution = FALSE}
str(pp)
```

```{r exo5-4-code-check}
grade_code()
```

### 

Using `groupe_by()`{.R} and `summarize()`{.R}:

- Show the number of males and females in the table (use the counter `n()`{.R})
- Show the average age per gender
- Show the average size per gender and institution
- Show the number of people from each country, sorted by descending population (`arrange()`{.R})

```{r exo5-5, exercise=TRUE, exercise.blanks = "___+", exercise.setup="exo5-4-solution", exercise.lines	= 20}
# - Show the number of males and females in the table (use the counter `n()`)
pp %>%
    ___ %>%
    ___
# - Show the average age per gender
pp %>%
    ___ %>%
    ___
# - Show the average size per gender and institution
pp %>%
    ___ %>%
    ___
# - Show the number of people from each country, sorted by descending population
pp %>%
    ___ %>%
    ___ %>%
    ___
```

```{r exo5-5-solution, exercise.reveal_solution = TRUE}
# - Show the number of males and females in the table (use the counter `n()`)
pp %>%
    group_by(gender) %>%
    summarize(count = n())
# - Show the average age per gender
pp %>%
    group_by(gender) %>%
    summarize(age = mean(age))
# - Show the average size per gender and institution
pp %>%
    group_by(gender, institution) %>%
    summarize(size = mean(size))
# - Show the number of people from each country, sorted by descending population
pp %>%
    group_by(origin) %>%
    summarize(count = n()) %>%
    arrange(desc(count))
```

```{r exo5-5-code-check}
grade_code()
```

###

Using `select()`{.R}, display:

- only the name and age columns
- all but the name column

```{r exo5-6, exercise=TRUE, exercise.blanks = "___+", exercise.setup="exo5-5-solution"}
# - only the name and age columns
pp ___
# - all but the name column
pp ___
```

```{r exo5-6-solution, exercise.reveal_solution = FALSE}
# - only the name and age columns
pp %>% select(c(name, age))
# - all but the name column
pp %>% select(-name)
```

```{r exo5-6-code-check}
grade_code()
```

### 

Using `filter()`{.R}, show data only for:

- Chinese people
- From institution ECL and UCBL
- People older than 22 
- People with a `e` in their name

```{r exo5-7, exercise=TRUE, exercise.blanks = "___+", exercise.setup="exo5-6-solution"}
# - Chinese people
pp ___
# - From institution ECL and UCBL
pp ___
# - People older than 22
pp ___
# - People with a `e` in their name
pp ___
```

```{r exo5-7-solution, exercise.reveal_solution = TRUE}
# - Chinese people
pp %>% filter(origin == "China")
# - From institution ECL and UCBL
pp %>% filter(institution %in% c("ECL", "UCBL"))
# - People older than 22
pp %>% filter(age > 22)
# - People with a `e` in their name
pp %>% filter(grepl("e", name))
```

```{r exo5-7-code-check}
grade_code()
```


## Exercise 6

Here we will see how to load many files at once with the `tidyverse` and how to perform some data wrangling.

### Loading the data

We will work with the files whose paths are stored in the vector `flist`. These files are all `.csv` files containing two columns and a header. We can use the fact that `read_csv()` accepts vectors as argument and read them all at once. 

- Read them all in a tidy tibble called `tib`.
- Make sure to add a column named `"file"` containing the list of filenames: look at the `id` parameter 
- Modify this `"file"` column so that it contains just the file name and not the full path – look at the `basename()` function.

```{r exo6-setup, include=FALSE}
flist <- tutor_examples("sample")
```

```{r exo6, exercise=TRUE, exercise.blanks = "___+"}
tib <- read_csv(___,           # what do we want to read?
               id = ___) %>%  # what is the name of the column containing the file names ?
        mutate(___)           # modify this column so that it contains just the file name and not the full path
```


```{r exo6-solution, exercise.reveal_solution = FALSE}
tib <- read_csv(flist, id = "file") %>% 
        mutate(file = basename(file))
```

```{r exo6-code-check}
grade_code()
```


### Operations on strings

We also want to get information from ou file names, such as the sample number, the temperature, the time, and the time unit. Use the function [`separate()`](https://tidyr.tidyverse.org/reference/separate.html) to split the `file` column into `sample`, `T`, `time` and `time_unit`. If applicable, make sure that the resulting columns are numeric by getting rid of the annoying characters.

```{r exo6-1-setup, include=FALSE}
flist <- tutor_examples("sample")
tib <- read_csv(flist, id = "file") %>%
    mutate(file = basename(file))
```

```{r exo6-1, exercise=TRUE, exercise.blanks = "___+"}
tib <- tib %>% 
    separate(col = ___, 
             into = ___, 
             convert = ___) %>% 
    mutate(sample = as.numeric(str_replace(___)),
           T = ___,
           time = ___,
           time_unit = ___
           )
tib
```

```{r exo6-1-solution, exercise.reveal_solution = TRUE}
tib <- tib %>% 
    separate(col = file, into = c("sample", "T", "time", "time_unit", NA), convert= TRUE) %>% 
    mutate(sample = as.numeric(str_replace(sample, "sample", "")),
           T = as.numeric(str_replace(T, "K", ""))
           )
tib
```

```{r exo6-1-code-check}
grade_code()
```

###

Now we want all times to be in the same unit. Using `mutate()` and `ifelse()`, convert the minutes in seconds, then get rid of the `time_unit` column.

```{r exo6-1bis-setup, include=FALSE}
flist <- tutor_examples("sample")
tib <- read_csv(flist, id = "file") %>%
    mutate(file = basename(file)) %>% 
    separate(col = file, into = c("sample", "T", "time", "time_unit", NA), convert= TRUE) %>% 
    mutate(sample = as.numeric(str_replace(sample, "sample", "")),
           T = as.numeric(str_replace(T, "K", ""))
           )
```

```{r exo6-1bis, exercise=TRUE, exercise.blanks = "___+"}
tib <- tib %>% 
    mutate(time = ifelse(test, yes, no)) %>% # convert minutes to seconds
    select(___) # get rid of the `time_unit` column
tib
```

```{r exo6-1bis-solution, exercise.reveal_solution = TRUE}
tib <- tib %>% 
    mutate(time = ifelse(time_unit=="min", time*60, time)) %>% 
    select(-time_unit)
tib
```

```{r exo6-1bis-code-check}
grade_code()
```

### Plotting data

Before going further, we want to take a look at our data using `ggplot`. Modify the following code so that a color is added as a function of the sample number, and the plots are gathered on a grid showing the time as a function of the temperature.

```{r exo6-2-setup, include=FALSE}
flist <- tutor_examples("sample")
tib <- read_csv(flist, id = "file") %>%
    mutate(file = basename(file)) %>% 
    separate(col = file, into = c("sample", "T", "time", "time_unit", NA), convert= TRUE) %>% 
    mutate(sample = as.numeric(str_replace(sample, "sample", "")),
           T = as.numeric(str_replace(T, "K", ""))
           ) %>% 
    mutate(time = ifelse(time_unit=="min", time*60, time)) %>% 
    select(-time_unit)
```

```{r exo6-2, exercise=TRUE, exercise.blanks = "___+"}
tib %>% 
    ggplot(aes(x = x, y = y, color = ___)) +
        geom_point() + 
        facet_grid(___ ~ ___)
```

```{r exo6-2-solution, exercise.reveal_solution = TRUE}
tib %>% 
    ggplot(aes(x = x, y = y, color = factor(sample))) +
        geom_point() + 
        facet_grid(time ~ T)
```

```{r exo6-2-code-check}
grade_code()
```

### Nesting data

We want to **nest** our data to be able to perform operations on them – like fitting them. Using the `nest()` function, nest the data so that we end up with only 2 columns: `file` and `data`.

```{r exo6-3-setup, include=FALSE}
flist <- tutor_examples("sample")
tib <- read_csv(flist, id = "file") %>%
    mutate(file = basename(file)) %>% 
    separate(col = file, into = c("sample", "T", "time", "time_unit", NA), convert= TRUE) %>% 
    mutate(sample = as.numeric(str_replace(sample, "sample", "")),
           T = as.numeric(str_replace(T, "K", ""))
           ) %>% 
    mutate(time = ifelse(time_unit=="min", time*60, time)) %>% 
    select(-time_unit)
```

```{r exo6-3, exercise=TRUE, exercise.blanks = "___+"}
tib <- tib %>% 
    nest(___)
tib
```

```{r exo6-3-hint}
# What are the columns to nest ? data = c(these_columns)
```

```{r exo6-3-solution, exercise.reveal_solution = FALSE}
tib <- tib %>% 
    nest(data = c(x, y))
tib
```

```{r exo6-3-code-check}
grade_code()
```


### Fitting all data

Now we can fit all our data at once with a linear model:

```{r exo6-4-setup, include=FALSE}
flist <- tutor_examples("sample")
tib <- read_csv(flist, id = "file") %>%
    mutate(file = basename(file)) %>% 
    separate(col = file, into = c("sample", "T", "time", "time_unit", NA), convert= TRUE) %>% 
    mutate(sample = as.numeric(str_replace(sample, "sample", "")),
           T = as.numeric(str_replace(T, "K", ""))
           ) %>% 
    mutate(time = ifelse(time_unit=="min", time*60, time)) %>% 
    select(-time_unit) %>%
    nest(data = c(x,y))
```

```{r exo6-4, exercise=TRUE, exercise.blanks = "___+"}
tib <- tib %>% 
    mutate(fit = map(data, ~lm(data=., x~y)))
tib
```

```{r exo6-4-solution, exercise.reveal_solution = TRUE}
tib <- tib %>%
    mutate(fit = map(data, ~ lm(data = ., x ~ y)))
tib
```

```{r exo6-4-code-check}
grade_code()
```