---
title: "Fits"
subtitle: "[Reproducible data treatment with R](http://lmi.cnrs.fr/r/fitting.html)"
author: "Colin Bousige"
output: 
    learnr::tutorial:
        progressive: true
        allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
# rmarkdown::run("fits.Rmd")
library(learnr)
library(tutor)
library(tidyverse)
gradethis::gradethis_setup()
```

## Exercise 1

Load the `tidyverse` library, then read the file `tutor_examples("exo_fit.txt")` into a `tibble`{.R} called `tofit`.

```{r exo1, exercise=TRUE, exercise.blanks = "___+"}
library(___)
tutor_examples("exo_fit.txt")
tofit <- ___
tofit
```

```{r exo1-solution, exercise.reveal_solution = FALSE}
tofit <- read_table(tutor_examples("exo_fit.txt"))
```

```{r exo1-code-check}
grade_code()
```

###

Plot all columns as a function of `x` to see your data. For this, use on-the-fly the function `pivot_longer()`{.R} to turn the tibble into a tidy one:

```{r exo1-1-setup, include=FALSE}
tofit <- read_table(tutor_examples("exo_fit.txt"))
```

```{r exo1-1, exercise=TRUE, exercise.blanks = "___+"}
tofit %>% 
    pivot_longer(cols = ___, # what column do we keep? cols = -this_column
                 values_to = ___, # name of the column gathering the values
                 names_to = ___ # names of the column gathering the variable names
                 ) %>% 
    ggplot(aes(x = ___, y = ___)) + # what columns contain the x and y values?
        geom_point() + # plot with points
        facet____(~ ___, scales = ___) # put the plots on a grid varying according to a column, with free scales
```

```{r exo1-1-solution, exercise.reveal_solution = FALSE}
tofit %>% 
    pivot_longer(cols = -x,
                 values_to = "y",
                 names_to = "variable"
                 ) %>% 
    ggplot(aes(x = x, y = y)) +
        geom_point() +
        facet_wrap(~variable, scales="free")
```

```{r exo1-1-code-check}
grade_code()
```


###

Using `lm()`{.R} or `nls()`{.R} fit each column as a function of `x` and display the "experimental" data and the fit on the same graph.
    + Tip: Take a look at the function `dnorm()`{.R} to define a Gaussian


```{r exo1-2-setup, include=FALSE}
tofit <- read_table(tutor_examples("exo_fit.txt"))
```

```{r exo1-2, exercise=TRUE, exercise.blanks = "___+", exercise.lines=31}
fit_y <- lm(data = ___, ___) # what is the dataframe? what do we fit against what?
predict(fit_y)
P1 <- ___ %>% # what dataframe are we working on?
    ggplot(aes(x = ___, y = ___)) + # what are the x and y values of the plot?
        geom____() + # plot the data to fit with points
        geom_line(___, col = ___) # how do we add the fitted values with a red line?
# # # # # # # # # # # # # # # # # # # # # # #
fit_z <- nls(data = ___, # the data
    ___, # the function
    start = list(___) # the list of starting parameters
)
P2 <- ___ %>% 
    ggplot(___) +
        ___
# # # # # # # # # # # # # # # # # # # # # # #
fit_w <- lm(___) # how can we fit w with a linear function?
P3 <- ___ %>% 
    ggplot(___) +
        ___
# # # # # # # # # # # # # # # # # # # # # # #
fit_u <- lm(___) # how can we fit u with a linear function?
P4 <- ___ %>% 
    ggplot(___) +
        ___
# # # # # # # # # # # # # # # # # # # # # # #
fit_s <- ___ # linear or non-linear fit?
P5 <- ___ %>% 
    ggplot(___) +
        ___
# # # # # # # # # # # # # # # # # # # # # # #
library(patchwork) # to gather the plots on a grid easily
P1 + P2 + P3 + P4 + P5 # plot on a grid
```


```{r exo1-2-solution, exercise.reveal_solution = TRUE}
fit_y <- lm(data = tofit, y ~ x)
P1 <- tofit %>% ggplot(aes(x = x, y = y)) +
    geom_point() +
    geom_line(aes(y = predict(fit_y)), col = "red")
# # # # # # # # # # # # # # # # # # # # # # # 
fit_z <- nls(data = tofit,
    z ~ A * dnorm(x, mean, sd),
    start = list(
        A = 8,
        mean = 4,
        sd = 1
    ))
P2 <- tofit %>% ggplot(aes(x = x, y = z)) +
    geom_point() +
    geom_line(aes(y = predict(fit_z)), col = "red")
# # # # # # # # # # # # # # # # # # # # # # #
fit_w <- lm(data = tofit, log(w) ~ x)
P3 <- tofit %>% ggplot(aes(x = x, y = log(w))) +
    geom_point() +
    geom_line(aes(y = predict(fit_w)), col = "red")
# # # # # # # # # # # # # # # # # # # # # # #
fit_u <- lm(data = tofit, exp(u) ~ x)
P4 <- tofit %>% ggplot(aes(x = x, y = exp(u))) +
    geom_point() +
    geom_line(aes(y = predict(fit_u)), col = "red")
# # # # # # # # # # # # # # # # # # # # # # #
fit_s <- lm(data = tofit, log(s) ~ x)
P5 <- tofit %>% ggplot(aes(x = x, y = log(s))) +
    geom_point() +
    geom_line(aes(y = predict(fit_s)), col = "red")
# # # # # # # # # # # # # # # # # # # # # # #
library(patchwork)
P1 + P2 + P3 + P4 + P5
```

```{r exo1-2-code-check}
grade_code()
```



