---
title: "Plots"
subtitle: "[Reproducible data treatment with R](http://lmi.cnrs.fr/r/plotting.html)"
author: "Colin Bousige"
output: 
    learnr::tutorial:
        progressive: true
        allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE, message=FALSE}
# rmarkdown::run("plots.Rmd")
library(learnr)
library(tidyverse)
library(tutor)
library(patchwork)
df <- read_table(tutor_examples("exo_fit.txt"))
```

## Exercise 1

We will work with the well known table `mtcars` included in R:

```{r}
str(mtcars)
```

### Basic stuff 1

Modify the following code to add a color depending on the `gear` column:
```{r q, exercise=TRUE, exercise.blanks = "___+"}
mtcars %>% # we work on the mtcars dataset, send it to ggplot
    ggplot(aes(x = wt, y = mpg))+ # define the x and y variables of the plot, and also the color
        geom_point() # plot with points
```

```{r q-solution, exercise.reveal_solution = TRUE}
mtcars %>%
    ggplot(aes(x = wt, y = mpg, color=gear)) +
    geom_point()
```

```{r q-code-check}
grade_code()
```

### Basic stuff 2

What happens if you use `factor(gear)` instead?
```{r q1, exercise=TRUE, exercise.blanks = "___+"}
mtcars %>% # we work on the mtcars dataset, send it to ggplot
    ggplot(aes(x = wt, y = mpg, color = gear))+ # define the x and y variables of the plot, and also the color
        geom_point() # plot with points
```

```{r q1-solution, exercise.reveal_solution = FALSE}
mtcars %>%
    ggplot(aes(x = wt, y = mpg, color = factor(gear))) +
    geom_point()
```

```{r q1-code-check}
grade_code()
```


### Faceting 1

Modify the following code to place each `carb` in a different facet. Also add a color, but remove the legend.
```{r q2, exercise=TRUE, exercise.blanks = "___+"}
mtcars %>% # we work on the mtcars dataset, send it to ggplot
    ggplot(aes(x = wt, y = mpg))+ # define the x and y variables of the plot, and also the color
        geom_point() +   # plot with points
        facet____(___) + # add a faceting
        theme(___)       # remove the legend
```

```{r q2-solution, exercise.reveal_solution = FALSE}
mtcars %>%
    ggplot(aes(x = wt, y = mpg, color = factor(gear)))+
    geom_point() +
    facet_wrap(~carb)+
    theme(legend.position = "none")
```

```{r q2-code-check}
grade_code()
```


### Faceting 2

Modify the following code to arrange `mpg` vs `wt` plots on a grid showing `gear` vs `carb`. Also add a color depending on `cyl`. Also, try adding a free `x` scale range, or a free `y` scale range, or free `x` and `y` scale ranges.
```{r q3, exercise=TRUE, exercise.blanks = "___+"}
mtcars %>% # we work on the mtcars dataset, send it to ggplot
    ggplot(aes(x = ___, y = ___))+ # define the x and y variables of the plot, and also the color
        geom_point() +   # plot with points
        facet____(___) # add a faceting
```

```{r q3-solution, exercise.reveal_solution = TRUE}
mtcars %>%
    ggplot(aes(x = wt, y = mpg, color = factor(cyl)))+
    geom_point() +
    facet_grid(paste("gear = ", gear) ~ paste("carb = ", carb), scales = "free")
```

```{r q3-code-check}
grade_code()
```



## Exercise 2

We will look at data loaded into `df`. 

```{r}
df
```

###

Using `ggplot`, plot `y` as a function of `x` with points and save it into `Py`:

```{r exo1-0, exercise=TRUE, exercise.blanks = "___+"}
Py <- df %>% 
    ___
Py
```

```{r exo1-0-solution, exercise.reveal_solution = TRUE}
Py <- df %>% 
    ggplot(aes(x = x, y = y)) + 
    geom_point()
Py
```

```{r exo1-0-code-check}
grade_code()
```


###

Add a straight line in `Py` resulting from a linear fit:

```{r exo1-1, exercise=TRUE, exercise.blanks = "___+"}
Py <- Py +
    geom_smooth(___)
Py
```

```{r exo1-1-solution, exercise.reveal_solution = TRUE, exercise.setup="exo1-0-solution"}
Py <- Py +
    geom_smooth(method = "lm")
Py
```

```{r exo1-1-code-check}
grade_code()
```

### 

Using `ggplot`, plot `z` as a function of `x` with a red line and save it into `Pz`:

```{r exo1-2, exercise=TRUE, exercise.blanks = "___+", exercise.setup="exo1-1-solution"}
Pz <- df %>% 
    ___
Pz
```

```{r exo1-2-solution, exercise.reveal_solution = TRUE}
Pz <- df %>% 
    ggplot(aes(x = x, y = z)) + 
    geom_line(color = "red")
Pz
```

```{r exo1-2-code-check}
grade_code()
```


### 

Using `ggplot`, plot a histogram of `w` with transparent blue bars surrounded by a red line, and save it into `Pw`. You can play with the number of bins too.

```{r exo1-3, exercise=TRUE, exercise.blanks = "___+", exercise.setup="exo1-2-solution"}
Pw <- df %>% 
    ___
Pw
```

```{r exo1-3-solution, exercise.reveal_solution = TRUE}
Pw <- df %>% 
    ggplot(aes(x = w)) + 
    geom_histogram(fill = "blue", alpha=.5, color="red", bins=50)
Pw
```

```{r exo1-3-code-check}
grade_code()
```


### 

Using `ggplot`, plot a density of `u` with a transparent blue area surrounded by a red line, and save it into `Pu`. Play with the `bw` parameter so that you see many peaks.

```{r exo1-4, exercise=TRUE, exercise.blanks = "___+", exercise.setup="exo1-3-solution"}
Pu <- df %>% 
    ___
Pu
```

```{r exo1-4-solution, exercise.reveal_solution = TRUE}
Pu <- df %>% 
    ggplot(aes(x = u)) + 
    geom_density(fill = "blue", alpha=.5, color="red", bw=.02)
Pu
```

```{r exo1-4-code-check}
grade_code()
```


### 

Using `patchwork`, gather the previous plots on a 2x2 grid.

```{r exo1-5, exercise=TRUE, exercise.blanks = "___+", exercise.setup="exo1-4-solution"}
Py
Pz
Pw
Pu
```

```{r exo1-5-solution, exercise.reveal_solution = TRUE}
library(patchwork)
Py+Pz+Pw+Pu
```

```{r exo1-5-code-check}
grade_code()
```


### 

Using `patchwork`, gather the previous plots on a grid with 3 plots in the 1st row, and one large plot in the 2nd row. Using `plot_annotation()`, add tags such as (a), (b)...

```{r exo1-6, exercise=TRUE, exercise.blanks = "___+", exercise.setup="exo1-4-solution"}
Py
Pz
Pw
Pu
```

```{r exo1-6-solution, exercise.reveal_solution = TRUE}
library(patchwork)
(Py + Pz + Pw) / Pu + plot_annotation(tag_levels = "a", tag_prefix="(", tag_suffix=")")
```

```{r exo1-6-code-check}
grade_code()
```
